Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes

### 2024-11-06 ms/ms.Rmd
ms: results - extrapolation outside mapped peatland

While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth > 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
For Ørskogfjelle we did not have such data.
Therefore, we first -- for both sites -- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
From the cross-validation we quantified *Root Mean Squared Error* and *Mean Absolute Error*, and we tested the strength of the correlation between predicted and observed peat depth.
Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.

@@

```{r modelMetrics, echo = FALSE, fig.cap = "Performance of models with different predictor configurations, evaluated via spatial cross-validation. Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error."}
knitr::include_graphics("figures/modelmetrics.pdf")
```

@@

>>>>

While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth > 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
For Ørskogfjelle we did not have such data.
Therefore, we first -- for both sites -- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
From the cross-validation we quantified *Root Mean Squared Error*.
Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
In this analysis, the model's prediction of peat depth is treated as a rank index of peatland likelihood.
Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.

@@

```{r modelMetrics, echo = FALSE, fig.cap = "Performance of peatland depth models with different predictor configurations, evaluated via spatial cross-validation. Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error."}
knitr::include_graphics("figures/modelmetrics.pdf")
```

@@

When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced marginally better RMSE than just assuming a constant \unit{30\,cm} depth (Fig. \@ref(fig:modelMetricsExtrapolation)).
Compared to our error estimates with the full dataset, RMSE at Skrimfjella was nearly unchanged, while at Ørskogfjellet it was lower outside of mapped peatland.
With the independent occurrence data at Skrimfjella, we found that neither the *terrain-only* nor *terrain and radiometric* configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).

```{r modelMetricsExtrapolation, echo = FALSE, fig.cap = "Performance of models that extrapolate from training data inside of mapped peatland to test data outside of mapped peatland, evaluated via spatial cross-validation. Parentheses denote the number of variables in each model, and point estimates are shown +/- their standard error."}
knitr::include_graphics("figures/modelmetrics-extrapolation.pdf")
```

### 2024-11-06 ms/ms.tex
ms: results - extrapolation outside mapped peatland

While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth \textgreater{} 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
For Ørskogfjelle we did not have such data.
Therefore, we first -- for both sites -- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
From the cross-validation we quantified \emph{Root Mean Squared Error} and \emph{Mean Absolute Error}, and we tested the strength of the correlation between predicted and observed peat depth.
Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.

@@

\begin{figure}
\centering
\includegraphics{figures/modelmetrics.pdf}
\caption{\label{fig:modelMetrics}Performance of models with different predictor configurations, evaluated via spatial cross-validation. Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error.}
\end{figure}

@@

>>>>

While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth \textgreater{} 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
For Ørskogfjelle we did not have such data.
Therefore, we first -- for both sites -- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
From the cross-validation we quantified \emph{Root Mean Squared Error}.
Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
In this analysis, the model's prediction of peat depth is treated as a rank index of peatland likelihood.
Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.

@@

\begin{figure}
\centering
\includegraphics{figures/modelmetrics.pdf}
\caption{\label{fig:modelMetrics}Performance of peatland depth models with different predictor configurations, evaluated via spatial cross-validation. Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error.}
\end{figure}

@@

When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced marginally better RMSE than just assuming a constant \unit{30\,cm} depth (Fig. \ref{fig:modelMetricsExtrapolation}).
Compared to our error estimates with the full dataset, RMSE at Skrimfjella was nearly unchanged, while at Ørskogfjellet it was lower outside of mapped peatland.
With the independent occurrence data at Skrimfjella, we found that neither the \emph{terrain-only} nor \emph{terrain and radiometric} configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).

\begin{figure}
\centering
\includegraphics{figures/modelmetrics-extrapolation.pdf}
\caption{\label{fig:modelMetricsExtrapolation}Performance of models that extrapolate from training data inside of mapped peatland to test data outside of mapped peatland, evaluated via spatial cross-validation. Parentheses denote the number of variables in each model, and point estimates are shown +/- their standard error.}
\end{figure}
